---
title: "Task 2 - Agglomerative hierarchical clustering"
author: "Max Settineri"
date: "2023-02-28"
output: 
  html_document:
    code_folding: hide
    toc: yes
---

## Introduction

This analysis explores stream chemistry data from the Santa Barbara Coastal drainage area. The code performs hierarchical clustering using complete linkages and creates a dendrogram showing clustering for water chemistry variables by stream site. The water chemistry variables explored in this analysis include data on specific conductivity, ammonium, nitrate, phosphorus, dissolved nitrogen and dissolved phosphorus concentrations. 

**Data Citation:** Citation: Santa Barbara Coastal LTER and J. Melack. 2019. SBC LTER: Land: Stream chemistry in the Santa Barbara Coastal drainage area, ongoing since 2000 ver 16. Environmental Data Initiative. https://doi.org/10.6073/pasta/67a558a24ceed9a0a5bf5e46ab841174.

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

## Data wrangling

```{r}
## reading in the data
streams <- read_csv(here('data/sbc_lter_registered_stream_chemistry.csv')) %>% 
  clean_names() %>% 
  na_if(-999) %>% ## changing -999 values to NA
  select(-tss_mgper_liter, -tpp_u_m, -tpn_u_m, -tpc_u_m, -timestamp_local) %>%  ## removing columns with more than 50% NA values
  group_by(site_code) %>% ## grouping by site
  summarize('Ammonium' = mean(nh4_u_m, na.rm = TRUE), ## renaming columns and using na.rm to remove rows with NA values
    'Nitrate' = mean(no3_u_m, na.rm = TRUE),
    'Phosphorous' = mean(po4_u_m, na.rm = TRUE),
    'Dissolved nitrogen' = mean(tdn_u_m, na.rm = TRUE),
    'Dissolved phosphorous' = mean(tdp_u_m, na.rm = TRUE),
    'Specific conductivity' = mean(spec_cond_u_spercm, na.rm = TRUE)) 
```

## Agglomerative hierarchical clustering

```{r}
## creating scaled version of the data for use in hierarchical clustering
streams_scaled <- streams %>% 
  select(-site_code) %>% ## removing site code since it's non numerical
  scale() ## scaling the data

## using site code as row names
rownames(streams_scaled) <- streams$site_code 

## creating matrix of euclidean distances
streams_dist <- dist(streams_scaled, method = 'euclidean')

## hierarchical clustering (complete linkage)
streams_hc_complete <- hclust(streams_dist, method = 'complete')

## maing a finalized dendrogram with ggdendro
ggdendrogram(streams_hc_complete) +
  theme_minimal() +
  labs(x = 'Stream Site', y = 'Euclidian Distance')

```

**Figure 1:** Depicts a dendrogram that shows relationships in water chemistry by stream site. The cluster dendrogram was produced using complete linkages. The dendrogram displays two main distinct clusters, with stream site `DV01` being isolated into its own third cluster.

## Summary

- Two main clusters are observed, with stream site `DV01` separated into its own cluster; this indicates that there are three total clusters of streams with similar water chemistry
  - `BC02`, `AT07`, `AB00`, `MC00`, `SP02`, `GV01` and `MC06` form the first main cluster
  - Stream sites `TO02`, `RG01`, `ON02`, `HO00` and `RS02` form the second main cluster
- The two most closely related stream sites are `HO00` and `RS02`
- More research should be conducted to determine which water chemistry variables contribute to these observed different clusters